Application of Generalized Additive Models for long-term and seasonal trend analysis of water quality: South San Francisco Bay case study

### 2020-04-29 manu_draft.Rmd
methods text draft done

The SFE Research Program of the US Geological Survey has been collecting water quality data at fixed locations in the bay since 1969 (Cloern and Schraga, 2016; Schraga et al.2017).  This dataset is one of the longest continuous monitoring datasets available in a North American estuary and includes at least monthly sampling at 37 fixed sampling locations along a 145 km longitudinal transect from the lower Sacramento River in the Delta to lower South Bay.  Water quality constituent measurements available for each station include salinity, temperature, light attenuation, chlorophyll-a, dissolved oxygen, suspended particulate matter, and dissolved inorganic nutrients. The current analysis focused on near-surface chlorophyll-a data collected biweekly to monthly along the South Bay axis extending from Central Bay (stations 18-23), South Bay (stations 24 - 32), and Lower South Bay (stations 34-36). Discrete chlorophyll concentrations at each station were determined by fluoremetric analysis with 90% acetone pigment extraction on GFF filters. Data collected between 1990-2017 were selected for analysis because it represented a suitable balance among three factors relevant to testing the statistical approaches, including sufficient length of record, consistent biweekly-monthly sampling, and a diverse set of stations covering the salinity gradient across multiple subembayments.  While sampling frequency varied somewhat over time or by station, all data were treated as unique time series within the statistical models (i.e., no spatial or temporal binning or averaging was done).

@@

To evaluate GAMs with SFE chloropyll data, we followed an approach that built on past work by the Chesapeake Bay Program (Murphy et al. 2019), including incorporation and adaptation of their hierarchical model structure to evaluate long-term time series.  These hierarchical models vary the functional form for time in relation to chlorophyll by adding smoothers and linear predictors with increasing flexibility in the GAM functions.  This is a powerful approach for time series analysis because separate components can be explicitly modelled with different smoothing terms and the results can be idendependently evaluated and further tested from the model output.  Varying degrees of model complexity also provides a balance between descriptive power and computational efficiency, where simpler models may be preferred in routine reporting or more complex models may be preferred for hypothesis testing.  We present four GAM structures, following a similar format as Murphy et al. (2019).  The notation for gam3, gam4, and gam5 are not used because describe models used in Murphy et al. (2019) that were developed for purposes specific to Chesapeake Bay.

@@

A challenge to optimizing GAMs for time series data with significant intra- and interannual variation is choosing an appropriate functional form (i.e., gam0, gam1, gam2, or gam6) and determining how much variation could be explained by the smoothers in each function.  The four models above provide tradeoffs in the functional forms that balance descriptive and computational efficiency.  However, the individual smoothing functions (i.e, $s_1()$, $s_2()$, $s_2^*()$ and $ti_1()$) require an explicit declaration of how much variation could be explained for the separate $year$ and $doy$ components.  This requires user input on the potential upper limit for the *k*-values that could be used during model fitting.  Initial testing by Murphy et al. (2019) showed that the default *k* values in the mgcv package were insufficient for describing the interannual variation in the response variable and specified the upper limit for the number of knots for the $s_2(year)$ smoother as the maximum between 10 or  $2 / 3 \times$ the number of years in the time series.   We follow the same approach herein for gam1 and gam2.  Further, the default *k*-values for $s_2()$ and $ti_1()$ were considered appropriate and not tested further. For the application to SFE chlorophyll data, gam6 was also added as an extension of gam1, where the total number of knots for the $year$ smoother, $s_2^*()$, was increased to 12 times the total number of years in the time series.  This approach potentially allows the smoother for $year$ to be completely unconstrained in how much variance is fit within and between years.

@@

The second hypothesis test evaluates changes in seasonal estimates between years using results from a fitted GAM.  This test provides both an estimate of a seasonal mean and its associated confidence interval and if the rate of change in the seasonal means has varied over a specific period interest.  For example,

* Secondary methods for trend analysis, estimate of seasonal values, hypothesis tests including mixed meta-analysis

## Sensitivity analysis

* Sensitivity analysis - do trend estimates or hypothesis test results change by model type?

>>>>

The SFE Research Program of the US Geological Survey has been collecting water quality data at fixed locations in the bay since 1969 (Cloern and Schraga, 2016; Schraga et al.2017).  This dataset is one of the longest continuous monitoring datasets available in a North American estuary and includes at least monthly sampling at 37 fixed sampling locations along a 145 km longitudinal transect from the lower Sacramento River in the Delta to lower South Bay.  Water quality constituent measurements available for each station include salinity, temperature, light attenuation, chlorophyll-a, dissolved oxygen, suspended particulate matter, and dissolved inorganic nutrients. The current analysis focused on near-surface chlorophyll-a data collected biweekly to monthly along the South Bay axis extending from Central Bay (stations 18-23), South Bay (stations 24-32), and Lower South Bay (stations 34-36). Discrete chlorophyll concentrations at each station were determined by fluoremetric analysis with 90% acetone pigment extraction on GFF filters. Data collected between 1990-2017 were selected for analysis because it represented a suitable balance among three factors relevant to testing the statistical approaches, including sufficient length of record, consistent biweekly-monthly sampling, and a diverse set of stations covering the salinity gradient across multiple subembayments.  While sampling frequency varied somewhat over time or by station, all data were treated as unique time series within the statistical models (i.e., no spatial or temporal binning or averaging was done).

@@

To evaluate GAMs with SFE chloropyll data, we followed an approach that built on past work by the Chesapeake Bay Program (Murphy et al. 2019), including incorporation and adaptation of their hierarchical model structure to evaluate long-term time series.  These hierarchical models vary the functional form for time in relation to chlorophyll by adding smoothers and linear predictors with increasing flexibility in the GAM functions.  This is a powerful approach for time series analysis because separate components can be explicitly modelled with different smoothing terms and the results can be idendependently evaluated and further tested from the model output.  Varying degrees of model complexity also provides a balance between descriptive power and computational efficiency, where simpler models may be preferred in routine reporting or more complex models may be preferred for hypothesis testing.  We present four GAM structures, following a similar format as Murphy et al. (2019).  The notation for gam3, gam4, and gam5 are not used because these describe models in Murphy et al. (2019) that were developed for purposes specific to Chesapeake Bay.

@@

A challenge for optimizing GAMs to time series data with significant intra- and interannual variation is choosing an appropriate functional form (i.e., gam0, gam1, gam2, or gam6) and determining how much variation could be explained by the smoothers in each function.  The four models above provide tradeoffs in the functional forms that balance descriptive and computational efficiency.  However, the individual smoothing functions (i.e, $s_1()$, $s_2()$, $s_2^*()$ and $ti_1()$) require an explicit declaration of how much variation could be explained for the separate $year$ and $doy$ components.  This requires user input on the potential upper limit for the *k*-values that could be used during model fitting.  Initial testing by Murphy et al. (2019) showed that the default *k* values in the mgcv package were insufficient for describing the interannual variation in the response variable and specified the upper limit for the number of knots for the $s_2(year)$ smoother as the maximum between 10 or  $2 / 3 \times$ the number of years in the time series.   We follow the same approach herein for gam1 and gam2.  Further, the default *k*-values for $s_2()$ and $ti_1()$ were considered appropriate and not tested further. For the application to SFE chlorophyll data, gam6 was also added as an extension of gam1, where the total number of knots for the $year$ smoother, $s_2^*()$, was increased to 12 times the total number of years in the time series.  This approach potentially allows the smoother for $year$ to be completely unconstrained in how much variance is fit within and between years.

@@

The second hypothesis test evaluates changes in seasonal estimates between years using results from a fitted GAM.  This test provides both an estimate of a seasonal mean, its associated confidence interval, and if the rate of change in the seasonal means has varied over a specific period of interest.  For example, changes that occur during late fall or early spring can be evaluated between years to determine if the seasonal averages differ and if the direction and magnitude of change is significant. This has implications for understanding potential changes in seasonal bloom phenology resulting from shifting system responses to nutrient inputs or changes in the latter.

Seasonal averages and confidence intervals were estimated from fitted GAMs by first creating a matrix of the predictors (i.e., `year`, `doy`) covering the seasonal periods across the time series.  For example, seasonal changes from late summer could be obtained from a prediction matrix covering all dates from Aug 1st to Sep. 30th for the period from 2000 to 2010.  The prediction matrix is then used to obtain fitted values from a GAM that describe the estimated values of chlorophyll that occurred during the periods of interest. Rather than the estimated values for the predictions, a linear predictor matrix is returned that provides a parameter vecotr at the supplied covariate values.  This predictor matrix is then used to estimate the mean seasonal value within a year and associated variance using the sum of predictions from the matrix.  As a result, each year has a mean value of chlorophyll with an associated confidence interval that describes the modelled expectation for the seasonal period.

Although the seasonal averages obtained from GAMs describe the expected values for each year, describing the direction and magnitude of the seasonal change across years may also be of interest.  Conventionally, trend analysis of changes in a response variable could be achieved with ordinary least squares regression, where the response variable is the expected value of chlorophyll and the predictor is year as a numeric value.  Regression analysis could be used to evaluate changes in the seasonal averages described above, although the fit from the regression would under-estimate the uncertainty in the model parameters by not accounting for the variance in the seasonal averages.  To better describe the trends in the seasonal averages, we use mixed effects regression meta-analysis to evaluate seasonal changes and their uncertainty over a finite set of years.  This analysis is an extension of conventional regression by accounting for different soures of uncertainty attributed to both variation of the true seasonal mean around the trend and variation of the estimated seasonal mean around the true seasonal mean:

\begin{equation}
chl_i = \beta_0 + \beta_1 year_i + b_i + \epsilon_i
(\#eq:mixmet)
\end{equation}

where the estimated value of chlorophyll at year $i$ for the predetermined season is a function of the predicted seasonal mean $\beta_0 +\beta_1 year_i$ plus deviation of the seasonal mean from the trend ($b_i$) and devation of the seasonal mean from the true mean ($S_i$).  Both the error terms are assumed to be normally distributed with mean zero and variance for $b_i$ equal to an unknown term and vraiance for $\epsilon_i$ equal to the squared standard error from the seasonal chloropyll averages.  The R package *mixmeta* was used to estimate trends in the seasonal averages over time (Sera et al. 2019).  Results appear similar to those from regression analysis (e.g., slope and intercept estimates), although parameter uncertainty is more accurately estimated with mixed meta-analysis.

## Model application and interpretation

Each of the four GAM structures were used to fit models to the chlorophyll time series at each of the nine stations from central to lower South Bay.  For each model and station, the fit was assessed using the standard summary statistics described above (i.e., GCV, AIC, and $R^2$) and relative significance of the individual smoothers terms.  Models were also evaluated using different transformations to determine which had improved fit for specific time series.  Secondary analyses evaluating trends with fitted GAM predictions were also compared between stations and model types, particularly to determine if and when significant results varied with alternative methods.  Temporal and spatial patterns within and between stations were also idenfied to better quanity changes in chlorophyll production throughout the period of record using GAMs to demonstrate improved descriptions relative to interpretations from the observed data.
